Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-02-05 paper/manuscript.tex
Working my way through section 5/6

Running free energy of binding calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations. The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes. The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs. Commercial tools can be expensive in some cases, but alternative tools are becoming available to allow relatively straightforward alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.

For relative free energy calculations, various graph topologies or maps of calculations are possible, and choices may depend on the target application. For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations should contain as many connections between compounds as affordable. However, when running calculations on hundreds of compounds a so called ‘star-map’ can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}. In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round. Alternatively, if the goal is to achieve the smallest possible error with minimal computational expense, certain graph topologies provide benefits~\cite{https://doi.org/10.1002/jcc.26095, https://doi.org/10.1021/acs.jcim.9b00528}

%
\subsection{Making predictions, understanding errors}
\label{subsec:predictions}
For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account.

It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}. Beyond this, several parameters provide useful indicators of performance. Results suggest MBAR errors greater than 0.3 kcal/mol for perturbations can be indicative of inaccuracy~\cite{perez-benito2019predicting}. In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure errors~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds. Selection bias, as described above, can be problematic in prospective applications as projects normally seek to focus only on the most potent compounds. Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.

In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved. Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study.

In a drug discovery setting it is inevitable to consider larger virtual libraries. In these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application. It is expected that future tools may allow efficient exploration of large virtual libraries. There are many accounts of success of alchemical FE calculations, the methods show good performance towards the goal of binding energy prediction. However, it is important to have realistic expectations.

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations. 
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Section~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Different types of simulations require different considerations for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\paragraph{How to deal with different binding modes?}
Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, as discussed in more detail in Section~\ref{sec:multiple_binding_modes}. Certain  functional groups may be particularly prone to this due to symmetries or near symmetries. One such issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series. The 180 degree flip of the ring may not occur enough during simulations to overcome bias due to the starting configuration. Another scenario may be equatorial and axial substituted saturated rings (e.g. cyclohexane derivatives). This situation may be addressed by explicitly modelling different binding modes of the same ligand and combining later computed free energy differences for different binding modes into a relative free energies of binding~\cite{kaus2015how}.
%
\paragraph{Have you considered sterioisomers and enationmers?}
Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly. For racemates, the relative abundance of each stereoisomer is normally not known. Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain. However, the modeling typically uses just the bioactive conformation that best fits the active site. Clearly this introduces potential for larger errors compared to experiment. Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust. Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer. Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
%
\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful. For examples, see HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
%
\paragraph{How to align congeneric series}
Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms. The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement. This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series. Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection, but there is a risk that the complex relaxes to an alternative metastable state.

@@

If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites. This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{10.1039/C6CP05680F} (REF). The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind. In this case a docking program can generate initial structures. Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, cheeseright2006molecular}.

>>>>

Running free energy of binding calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations. The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes. The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs. Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace}.

For relative free energy calculations, various graph topologies or maps of calculations are possible, and choices may depend on the target application. For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations should contain as many connections between compounds as affordable. However, when running calculations on hundreds of compounds a so called \emph{star-map} (see Fig.~\ref{fig:fig_types_of_networks} \textbf{A}) can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}. In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round. Alternatively, if the goal is to achieve the smallest possible error with minimal computational expense, certain graph topologies provide benefits~\cite{https://doi.org/10.1002/jcc.26095, https://doi.org/10.1021/acs.jcim.9b00528}

\subsection{Making predictions, understanding errors}
\label{subsec:predictions}
For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account.

It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}. Beyond this, several parameters provide useful indicators of performance. For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}. Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}). Selection bias, as described above, can be problematic in prospective applications as projects normally seek to focus only on the most potent compounds. Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}. For a more detailed discussion on checking the robsustness of your alchemcial free energy calculation see also Section~\ref{sec:are-they-good}.

In summary, the successful use of alchemical calculations not only in, but particularly for drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved. Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study. 
 There are many accounts of success of alchemical calculations, the methods show good performance towards the goal of binding energy prediction. However, it is important to have realistic expectations.

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Section~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\paragraph{How to deal with different binding modes?}
Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, as discussed in more detail in Section~\ref{sec:multiple_binding_modes}. Certain  functional groups may be particularly prone to this due to symmetries or near symmetries. One such issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series. The 180 degree flip of the ring may not occur enough during simulations to overcome bias due to the starting configuration. Another scenario may be equatorial and axial substituted saturated rings (e.g. cyclohexane derivatives). This situation may be addressed by explicitly modelling different binding modes of the same ligand and combining later computed free energy differences for different binding modes into a relative free energies of binding~\cite{kaus2015how}.
%
\paragraph{Have you considered sterioisomers and enationmers?}
Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly. For racemates, the relative abundance of each stereoisomer is normally not known. Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain. However, the modeling typically uses just the bioactive conformation that best fits the active site. Clearly this introduces potential for larger errors compared to experiment. Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust. Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer. Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
%
\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful. For examples, see HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.

\paragraph{How to align congeneric series}
Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms. The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement. This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series. Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection, but there is a risk that the complex relaxes to an alternative metastable state.

@@

If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites. This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{evoli2016multiple}. The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind. In this case a docking program can generate initial structures. Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, cheeseright2006molecular}.
